Comparing variations in sediment transport capacity between subglacial and subaerial systems

### 2023-03-23 Delaney2023.tex
more small changes

\item  Greater variability in water velocity in subglacial systems causes sediment transport capacity to vary more compared to subaerial systems.

@@

\begin{abstract} % 150 words
Sediment transport capacity depends mainly on the shear stress or velocity of water flowing through a channel and the channel width over which to mobilize sediment.
In subaerial channels, changing water discharge can be accommodated by rapid changes in the width of the channel over which water flows, the water depth, and the water velocity.
In subglacial channels, however, water flows through a conduit whose size evolves  slowly, because the water is pressurized by the glacier ice above.
As a result, variations in water discharge may modify the water velocity, but not the channel's width.
Here, parameterizations of subglacial and subaerial channels indicate that sediment transport capacity varies more in subglacial systems compared with subaerial systems, across a wide range of channel geometries and flow conditions.
Large variations in subglacial water sediment transport capacity may impact sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.

@@

\section*{Plain language summary} % 200 words
In both subglacial and subaerial channels, sediment transport capacity depends mainly on 1) the channel width over which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom; the shear stress depends largely on the water velocity.
In subaerial channels, changing water discharge will  change the water depth,  the channel width (which dictates the distance  over which sediment transport  occurs), and the  water velocity (which controls the sediment transport capacity).
 In subglacial environments, however, on times scales of hours to weeks, changing water discharge will only alter  water velocity, not channel area, because the subglacial conduit maintains a largely fixed geometry.
The fixed conduit geometry also means that the channel width over which the sediment is  mobilized also does not respond to changing water discharge.
Parameterizations of these processes show that sediment transport capacity varies far more in subglacial systems compared to subaerial systems, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
The increased variability in sediment transport capacity in subglacial systems, compared to subaerial ones, has several important implications for sediment transport in glacierized catchments and the interpretation of sediment discharge records.

@@

  Glaciers expel large quantities of sediment from their termini \cite{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>[]{zhang2022}.
  Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
  In some of these regions, increased water discharge and glacier melt have been attributed to the greater sediment transport \cite{bendixen2017,costa2017,li2021}.
  In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \cite{delaney2020,vergara2022}.
  Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
  Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

@@

  Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the subaerial system (Figure~\ref{fig:model_outs}).
  Water velocity in the glacier system vary by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the subaerial system the water velocity only varies by $0.11$ (Figure~\ref{fig:model_outs}\,e).
  Shear stress at the glacier bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the subaerial system varies by roughly $0.21$ (Figure~\ref{fig:model_outs}\,f). The increased variability in shear stress relative to water velocity between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}.

@@

  Additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to subaerial systems \cite{jansson2005,ganti2016}. 
  For instance, sediment transport capacity from a glacier reacts to the ice thickness, controlling the closure of the channel, and the surface slope of the glacier, controlling water velocity \cite<Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,delaney2022,stevens2022}, in addition to water discharge.
  Conversely, sediment transport in most subaerial systems increases with water discharge and bed slope, with the bed slope probably being more stable over yearly to century-scale time periods \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,whipple1999,wong2006,wickert2019}.

  Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \cite<e.g.>[]{koppes2009a,willenbring2016,mariotti2021}.
  The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels.
  Yet, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to subaerial systems \cite{tofelde2021}.
  This higher threshold of climatic perturbation may persist in addition to the stochastic nature of erosion and deposition in fluvial environments \cite{castletort2003,jerolmack2010,romans2016}.
  For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \cite{delaney2018}.
  Such a large quantity of sediment transported at lower water discharges may occur in part because sediment transport capacity responds the size of the glacier conduit, in addition to water discharge (Section~\ref{sect:sub_mode}).

>>>>

\item  Greater variability in water velocity in subglacial systems causes sediment transport capacity to vary more than subaerial systems.

@@

\begin{abstract} % 150 words
Sediment transport capacity depends mainly on the shear stress or velocity of water flowing through a channel and the channel width over which to mobilize sediment.
In subaerial channels, changing water discharge can be accommodated by rapid changes in the channel width over which water flows, the water depth, and the water velocity.
In subglacial channels, however, water flows through a conduit whose size evolves  slowly, because the water is pressurized by the glacier ice above.
As a result, variations in water discharge may modify the water velocity, but not the channel's width.
Here, parameterizations of subglacial and subaerial channels indicate that sediment transport capacity varies more in subglacial systems compared with subaerial systems, across a wide range of channel geometries and flow conditions.
Large variations in subglacial water sediment transport capacity may impact sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.

@@

\section*{Plain language summary} % 200 words
In both subglacial and subaerial channels, sediment transport capacity depends mainly on 1) the channel width over which the sediment is mobilized  and 2) the shear stress applied by flowing water on sediment at the channel bottom; the shear stress depends mainly on the water velocity.
In subaerial channels, changing water discharge will  change the water depth,  the channel width (which dictates the distance  over which sediment transport  occurs), and the  water velocity (which controls the sediment transport capacity).
 In subglacial environments, however, on times scales of hours to weeks, changing water discharge will only alter  water velocity, not channel area, because the subglacial conduit maintains a largely fixed geometry.
The fixed conduit geometry also means that the channel width over which the sediment is  mobilized also does not respond to changing water discharge.
Parameterizations of these processes show that sediment transport capacity varies far more in subglacial systems compared to subaerial systems, even though changing channel width can accommodate some sediment transport variability in subaerial channels.
The increased variability in sediment transport capacity in subglacial systems, compared to subaerial ones, has several important implications for sediment transport in glacierized catchments and the interpretation of sediment discharge records.

@@

  Glaciers expel large quantities of sediment from their termini \cite{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>[]{zhang2022}.
  Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
  In some of these regions, increased water discharge and glacier melt have been attributed to greater sediment transport \cite{bendixen2017,costa2017,li2021}.
  In other cases, enhanced access of meltwater to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \cite{delaney2020,vergara2022}.
  Observed changes to sediment transport in glacierized catchments create an imperative to examine the processes controlling sediment discharge here. 
  Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

@@

  Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the subaerial system (Figure~\ref{fig:model_outs}).
  Water velocity in the glacier system varies by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the subaerial system the water velocity only varies by $0.11$ (Figure~\ref{fig:model_outs}\,e).
  Shear stress at the glacier bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the subaerial system varies by roughly $0.21$ (Figure~\ref{fig:model_outs}\,f). The increased variability in shear stress relative to water velocity between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}.

@@

  Additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to subaerial systems \cite{jansson2005,ganti2016}. 
  For instance, sediment transport capacity from a glacier reacts to the ice thickness, controlling the closure of the channel, and the surface slope of the glacier, controlling water velocity \cite<Section~\ref{sect:sub_mode}; >{rothlisberger1972,shreve1972,delaney2022,stevens2022}, in addition to water discharge.
  Conversely, sediment transport in most subaerial systems increases with water discharge and bed slope, and the bed slope probably remains relatively stable over yearly to century-scale time periods \cite<Section~\ref{sect:fluv}; e.g.>{muller1968,whipple1999,wong2006,wickert2019}.

  Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \, cite<e.g.>[]{koppes2009a,willenbring2016,mariotti2021}.
  The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels.
  Yet, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to subaerial systems \cite{tofelde2021}.
  This higher threshold of climatic perturbation may persist in addition to the stochastic nature of erosion and deposition in fluvial environments \cite{castletort2003,jerolmack2010,romans2016}.
  For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \cite{delaney2018}.
  Such a large quantity of sediment transported at lower water discharges may occur in part because sediment transport capacity responds to the size of the glacier conduit, in addition to water discharge (Section~\ref{sect:sub_mode}).

### 2023-03-23 theory.tex
more small changes

\paragraph{Key findings:}
\begin{enumerate}
\item Water discharge variations in subglacial systems are accommodated by water velocity, not channel size, given the slow evolution of channels.

@@

  The variance $ \chi$ of a variable $\Gamma$ is given by 
  \begin{equation}
    \label{eq:var}
    \chi(\Gamma) \,=\, 1 - \frac{\mathrm{min}(\Gamma)}{\mathrm{max}(\Gamma)}\,\,.
  \end{equation}
  % 
  \noindent Below we examine variables, $\Gamma$, as water discharge ($Q_w$) water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and integrated shear stress ($\tau_{tg}$, $\tau_{tf}$) with the purpose of evaluating total shear stress across the glacier bed.

@@

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.7\linewidth]{model_outputs.pdf}
      \caption{Response of channel width (a), hydraulic diameter (b),  Reynolds number  (c),  water velocity (d), shear-stress (e), and width-integrated shear-stress (f)  to different water discharge values. The glacial system is in blue and the subaerial system is in green.  Horizontal bars in d, e, and f represent the range of values in the  output. Gray space represents water discharge below the mean value. }
      \label{fig:model_outs}
    \end{figure}
  \end{center}

  Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the subaerial system (Figure~\ref{fig:model_outs}).
  Water velocity in the glacier system vary by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the subaerial system the water velocity only varies by $0.11$ (Figure~\ref{fig:model_outs}\,e).
  Shear stress at the glacier bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the subaerial system varies by roughly $0.21$ (Figure~\ref{fig:model_outs}\,f). The increased variability in shear stress relative to water velocity between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}.

@@

@@

  Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \citep{gimbert2016,perolo2018}.
  The rapid increase and decreases in sediment transport capacity, for example over a diurnal cycle, in a subglacial system may compound the variability in sediment transport already present in subaerial systems \citep{williams1989,jerolmack2010}.
  As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \citep[e.g.][]{richards2003,swift2021}.
  Aside from the sporadic nature of these events, the observed flushings could result from increases in water velocity resulting from reduced channel size, not water discharge, which increase sediment transport capacity.

>>>>

\paragraph{Key findings:}
\begin{enumerate}
\item Water discharge variations in subglacial systems are accommodated by water velocity, not channel size, given the channel's slow evolution.

@@

  The variance $ \chi$ of a variable $\Gamma$ is given by 
  \begin{equation}
    \label{eq:var}
    \chi(\Gamma) \,=\, 1 - \frac{\mathrm{min}(\Gamma)}{\mathrm{max}(\Gamma)}\,\,.
  \end{equation}
  % 
  \noindent The results below examine variables $\Gamma$, as water discharge ($Q_w$) water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and integrated shear stress ($\tau_{tg}$, $\tau_{tf}$) with the purpose of evaluating total shear stress across the glacier bed.

@@

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.7\linewidth]{model_outputs.pdf}
      \caption{Response of channel width (a), hydraulic diameter (b),  Reynolds number  (c),  water velocity (d), shear-stress (e), and width-integrated shear-stress (f)  to different water discharge values. The glacial system is in blue and the subaerial system is in green.  Horizontal bars in d, e, and f represent the range of values in the  output. Gray space represents water discharge below the mean value.}
      \label{fig:model_outs}
    \end{figure}
  \end{center}

  Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the subaerial system (Figure~\ref{fig:model_outs}).
  Water velocity in the glacier system varies by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the subaerial system the water velocity only varies by $0.11$ (Figure~\ref{fig:model_outs}\,e).
  Shear stress at the glacier bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the subaerial system varies by roughly $0.21$ (Figure~\ref{fig:model_outs}\,f). The increased variability in shear stress relative to water velocity between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}.

@@

  Both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous \citep[Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}][]{yager2018}. 
  The subaerial parameterization assumes that sediment is transported across all channel widths, as opposed to large flows such as floods when much sediment transport occurs \citep{wolman1960}.
  As a result, the variability in velocity, shear-stress, and width integrated shear stress in the subaerial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

@@

  Large variations in sediment transport capacity below glaciers may result in sediment mobilization and deposition processes in close spatial and temporal proximity to each other below the glacier \citep{gimbert2016,perolo2018}.
  The rapid increase and decreases in sediment transport capacity, for example over a diurnal cycle, in a subglacial system may compound the variability in sediment transport already present in subaerial systems \citep{williams1989,jerolmack2010}.
  As a result, variations in sediment transport capacity could be responsible for the fluctuations in sediment discharge (``flushings'') from glaciers that are not attributed to variations in water discharge \citep[e.g.][]{richards2003,swift2021}.
  Aside from the sporadic nature of these events, the observed "flushing" could result from increases in water velocity resulting from reduced channel size, not water discharge, which increase sediment transport capacity.
